Tracing the cultural evolution of meaning at the semantics-pragmatics interface

### 2017-03-15 paper/paper-v3.tex
minor additions new setup draft

\begin{abstract}
  According to standard linguistic theory, the meaning of an utterance is the product of 
  conventional semantic meaning and general pragmatic reasoning applied
  to the context of utterance. This implies that models of cultural evolution of meaning
  should likewise take into consideration that observable language use is a complex interaction
  of semantic representations and pragmatic use. To this end, we present a game theoretic model
  of cultural language evolution where communicative pressures work on abstract semantic
  representations and pragmatic patterns of use. Our model traces two evolutionary forces and
  their interaction: (i) fitness-based pressure towards communicative efficiency and (ii)
  systematic transmission perturbations when linguistic knowledge is transferred from one agent
  to another.  We illustrate the model based on a case study showing that learning biases that favor simple semantic
  representations can prevent the lexicalization of pragmatic inferences as long as mutual reasoning counteracts a potential loss in expressivity from such semantics. 
\end{abstract}

\section{Introduction}\label{sec:introduction}
What is conveyed usually goes beyond what is said. A request for a blanket can be politely
veiled by uttering ``I'm cold.'' The temporal succession of events can be communicated by the
order in which conjuncts appear as in ``I traveled to Paris and got married.'' An invitation
can be declined by saying ``I have to work.'' An influential explanation of the relation
between the literal meaning of expressions and what they are intended and interpreted to convey
is due to \citet{grice:1975}, characterizing pragmatic use and interpretation as a process of
mutual reasoning about rational language use. For instance, under the assumption that the
speaker is cooperative and relevant, ``I have to work'' may be interpreted as providing a
reason why the speaker will not be able to accept the invitation, going beyond its literal
meaning. Some of these enrichments are rather \emph{ad hoc}. Others show striking regularities,
such as the use of ability questions for polite requests (``Could you please \dots?''), or
certain enrichments of lexical meanings such as \emph{and} to mean \emph{and then}.

A particularly productive and well studied class of pragmatic enrichments are scalar
implicatures
\citep{horn:1984,Hirschberg1985:A-Theory-of-Sca,LevinsonPragmatics1983,Geurts2010:Quantity-Implic}. Usually,
the utterance of a sentence like ``I own some of Johnny Cash's albums'' will be taken to mean
that the speaker does not own all of them. This is because, if the speaker had them all, he
could have used the stronger word \emph{all} instead of \emph{some} in his utterance, 
thereby making a more informative statement. Scalar implicatures, especially the
inference from \emph{some} to \emph{some but not all}, have been studied extensively, both
theoretically
\citep[e.g.,][]{Sauerland2004:Scalar-Implicat,ChierchiaFox2008:The-Grammatical,Rooyvan-RooijJagerde-Jager2012:Explaining-Quan}
as well as experimentally
\citep[e.g.,][]{BottNoveck2004:Some-Utterances,huang+snedeker:2009,GrodnerKlein2010:Some-and-Possib,GoodmanStuhlmuller2013:Knowledge-and-I,DegenTanenhaus2012:Processing-Scal}. While
there is much dispute in this domain about many interesting details, a position endorsed by a
clear majority is that a word like \emph{some} is underspecified to mean \emph{some and maybe
  all} and that the enrichment to \emph{some but not all} is part of some regular 
process with roots in pragmatics.

@@

To fill this gap, we spell out a model of the co-evolution of conventional meaning and
pragmatic reasoning types. The objects of replication and selection are pairs of lexical
meanings and general types of pragmatic behavior, which we represent using state-of-the-art
probabilistic cognitive models of pragmatic language use
\citep{frank+goodman:2012,FrankeJager2015:Probabilistic-p,GoodmanFrank2016:Pragmatic-Langu}. Replication
and selection are described by the \emph{replicator mutator dynamic}, a general and established
model of evolutionary change in large and homogeneous populations
\citep{Hofbauer1985:The-Selection-M,nowak+etal:2000,NowakKomarova2001:Evolution-of-Un,hofbauer+sigmund:2003,Nowak2006:Evolutionary-Dy}. The
approach allows us to study the interaction between (i) evolutionary pressure towards
communicative efficiency and (ii) possible infidelity in the transmission of linguistic
knowledge, caused by factors such as inductive learning biases and general aspects of learnability. Considering
transmission of linguistic knowledge is important because neither semantic meanings nor
pragmatic usage patterns are directly observable. Instead, language learners have to infer these
unobservables from the observable behavior in which they result. We formalize this 
process as a form of Bayesian inference. Our approach thereby contains a well-understood model
of iterated Bayesian learning \citep{griffiths+kalish:2007} as a special case, but combines it
with functional selection, here formalized as the most versatile dynamic from evolutionary game
theory; the replicator dynamic
\citep{TaylorJonker1978:Evolutionary-St}. Section~\ref{sec:model} introduces this model.

@@

An elegant formal approach to capturing the interaction between expressivity and learnability
is the \emph{replicator mutator dynamic}
\citep{Hofbauer1985:The-Selection-M,nowak+etal:2000,NowakKomarova2001:Evolution-of-Un,hofbauer+sigmund:2003,Nowak2006:Evolutionary-Dy}. In
its simplest, discrete-time formulation, the RMD defines the frequency $x'_i$ of each type $i$
in a population at the next time step as a function of: (i) the frequency $x_i$ of each type
$i$ before the step, (ii) the fitness $f_i$ of each type $i$ before the step, and (iii) the
probability $Q_{ji}$ that an agent who wants to imitate, adopt, or learn the type of an agent
with type $j$ ends up acquiring type $i$:
\begin{align}
  \label{eq:RMD_discrete}
  x'_i = \sum_j Q_{ji} \frac{x_jf_j}{\sum_k x_k f_k}\,.
\end{align}
The RMD consists of two components: fitness-based selection and transmission biases. This
becomes most transparent when we consider an equivalent formulation in terms of a step-wise
application of the discrete-time replicator dynamic \citep{TaylorJonker1978:Evolutionary-St} on the initial population vector $\vec{x}$
and its subsequent multiplication with a mutation matrix $Q$:

@@

Where our goal is an application of this dynamic to the case of co-evolution of semantic
meaning and pragmatic use, we need to fix what the relevant types are, how fitness is measured
and how the mutation matrix is computed. These issues will be addressed, one by one, in the
following.

@@

Lexica codify the truth-conditions of expressions. A convenient way to represent lexica is by
$(\card{\States}, \card{\Messgs})$-Boolean matrices, where $\States$ is a set of states
(meanings) and $M$ a set of messages (forms available in the language). For example, suppose
that there are two relevant world states $\States = \set{\ssome, \sall}$. In state $\ssome$
Chris owns some but not all of Johnny Cash's albums while in $\sall$ Chris owns them
all. Suppose that there are two messages $\Messgs = \set{\msome, \mall}$ where $\msome$ is
short for a sentence like \emph{Chris owns some of Johnny Cash's albums} and $\mall$ for the
same sentence with \emph{some} replaced by \emph{all}. \tb{I left $\snone$/$\mnone$ out because they are not relevant to the strengthening of $\msome$, but that means that it needs to be introduced a bit later. Feel free to add it if you think that's better, exposition-wise.} 
Lexica for this case would assign a
truth value for each state-message pair. The following two lexica exemplify the distinction
between a lexicalized upper-bound for \emph{some} in $\Lbound$ and the widely assumed logical
semantics with only a lower-bound in $\Llack$.

@@

We distinguish between two kinds of pragmatic behavior. {\em Literal interlocutors} produce and
interpret messages literally, being guided only by their lexica. {\em Pragmatic interlocutors}
instead engage in mutual reasoning to inform their choices. Recent probabilistic models of
rational language use
\citep{franke:2009,frank+goodman:2012,FrankeJager2015:Probabilistic-p,GoodmanFrank2016:Pragmatic-Langu}
capture different types of pragmatic behavior in a reasoning hierarchy. The hierarchy's bottom,
level $0$, corresponds to literal language use. Pragmatic language users of level $n + 1$
act (approximately) rational with respect to level-$n$ behavior of their
interlocutors. (\ref{h:level0}) and (\ref{s:level0}) define probabilistic behavior of literal
hearers and speakers respectively: \mf{is it a problem that $S$ denotes speakers and the set of
states?} \tb{I think that's fine. $S_n(\cdot \mid \cdot)$ is not used beyond this section and the set $S$ only by mention of its members. Otherwise we can, e.g., relabel $M$ as $F$(orms) and use $M$(eanings) instead of $S$(tates). I'm fine with either alternative.}
\begin{flalign}
&H_{0}(s \mid m;L) \propto pr(s) L_{sm} \label{h:level0}\\
&S_{0}(m \mid s;L) \propto \exp(\lambda \; L_{sm}) \label{s:level0}
\end{flalign}

According to (\ref{h:level0}), a literal hearer's interpretation of a message $m$ as a state
$s$ depends on her lexicon and her prior over states, $pr \in \Delta(S)$. For simplicity, in
the following this prior is assumed to be uniform. A literal interpreter with lexicon $\Lbound$
from above would assign $\ssome$ a probability of $H_0(\ssome \mid \msome; \Lbound) = 1$ after
hearing $\msome$, while a literal interpreter with lexicon $\Llack$ would assign $\ssome$
probability $H_0(\ssome \mid \msome; \Llack) = 0.5$. As usual in probabilistic pragmatics models, speaker behavior is regulated by a soft-max
parameter $\lambda$, $\lambda \geq 1$ \citep{luce:1959,sutton+barto:1998}. As $\lambda$
increases, choices made in production are more rational in that higher values lead to behavior
that is increasingly in line with expected utility maximization. Expected utility of a message
$\messg$ in state $\state$ for a level $n+1$ speaker is here defined as $H_{n}(s|m;L)$, the probability that the hearer will assign to or choose the correct meaning. Put differently, how likely the hearer is thought to infer $\state$ when sent $\messg$. The formulation
in (\ref{s:level0}) also allows false messages to be sent with a small positive probability in
analogy to the definition of pragmatic speakers in probabilistic pragmatic models. This is
necessary to guarantee a mutation matrix with only positive entries (see below). If $\lambda = 1$ a literal
speaker with either lexicon $\Lbound$ or $\Llack$ produces $\msome$ in $\ssome$ with probability $S_{0}(\msome \mid
\ssome;L_{\text{bound},\text{lack}}) \approx .73$. The probability of producing $\msome$ in $\sall$ is $S_{0}(\msome \mid
\sall;\Lbound) \approx .27$ for $\Lbound$ and $S_{0}(\msome \mid
\sall;\Llack) \approx .5$ for $\Llack$. By contrast, if $\lambda = 20$, a literal speaker with either lexicon produces $\msome$ in $\ssome$ with probability $S_{0}(\msome \mid
\ssome;L_{\text{bound},\text{lack}}) \approx 1$. But the probability of producing $\msome$ in $\sall$ is $S_{0}(\msome \mid
\sall;\Lbound) \approx 1$ for $\Lbound$ whereas it is still $S_{0}(\msome \mid \sall;\Llack) \approx .5$ for $\Llack$.

Pragmatic behavior of level-$n+1$ is similar to its literal counterparts but uses the
interpretation or production behavior of a level-$n$ player instead of the lexical meaning:
\begin{flalign}
&H_{n+1}(s|m;L) \propto pr(s) S_{n}(m|s;L) \label{h:leveln}\\
&S_{n+1}(m|s;L) \propto  \exp(\lambda \; H_{n}(s|m;L)) \label{s:leveln}
\end{flalign}

Particularly important to our purpose is the fact that players of a level greater than $0$ using $\Llack$ can {\em pragmatically} associate $\msome$ with $\ssome$ over $\sall$, in contrast to their literal counterparts. Intuitively, this is so because hearers reason a speaker to convey $\sall$ with $\mall$, as this has a higher chance of leading to mutual understanding than using underspecified $\msome$. Therefore they will expect $\msome$ to be more strongly associated with $\ssome$ than with $\sall$ as it is already possible to clearly convey the latter state with a different message. The converse reasoning pattern holds for the speaker.

@@

Languages are shaped not only by functionalist forces towards greater expressivity. Another
important factor is the fidelity by which language is transmitted. Among others, linguistic
production can be prone to errors, states or messages may be perceived incorrectly, and
multiple languages may be compatible with the data learners are exposed to. These sources of
uncertainty introduce variation in their transmission from one generation to the next. In particular, learning biases in the iterated transmission process can influence language evolution
substantially.

@@

Following \citet{griffiths+kalish:2007} we model language acquisition as a process of Bayesian
inference in which learners combine the likelihood of a type producing the witnessed learning
input with prior inductive biases. Experimental and mathematical results on iterated learning
suggest that the outcome of this process reflects learners' inductive biases
\citep[e.g.,][]{kirby+etal:2014}. In a Bayesian setting these biases can be codified in a prior
$P \in \Delta(T)$, which reflects the amount of data a learner requires to faithfully acquire
the language of the teacher \citep[cf.][450]{griffiths+kalish:2007}. The extent of the prior's
influence has been shown to heavily depend on the learning strategy assumed to underly the
inference process. On the one hand, early simulation results suggested that weak biases could
be magnified by exposing learners to only small data samples \citep[e.g., in][]{brighton:2002}. On the other, \citeposs{griffiths+kalish:2007} mathematical
characterization showed that iterated learning converged to the prior, i.e., the resulting
distribution over languages corresponds to the learners' prior distribution and is not
influenced by the amount of input given to them. This difference in predictions can be traced
back to differences in the selection of hypotheses from the posterior. Griffith \& Kalish's
convergence to the prior holds for learners that sample from the posterior. More deterministic
strategies such as the adoption of the type with the highest posterior probability, so-called
{\it maximum a posterior estimation} (MAP), increase the influence of both the prior and the
data \citep{griffiths+kalish:2007,kirby+etal:2007}. In the following, we use a parameter
$l\ge1$ to modulate between posterior sampling and the MAP strategy. When $l = 1$ learners
sample from the posterior. The learners' propensity to maximize the posterior grows as $l$
increases.

@@

Expressivity and learnability are central to the cultural evolution of language. These
components can be modelled, respectively, as replication based on a measure of fitness in terms
of communicative efficiency and iterated Bayesian learning. Their interaction is described by
the discrete time replicator mutator dynamics in (\ref{eq:RMD_discrete}), repeated here:
\begin{align*}
  x'_i = \sum_j Q_{ji} \frac{x_jf_j}{\sum_k x_k f_k}\,.
\end{align*}
This equation defines the frequency $x'_i$ of type $i$ at the next time step, based on its
frequency $x_i$ before the step, its fitness $f_i$ and the probability that a learner infers
$i$ when observing the behavior of a type-$j$ agent. Fitness-based selection can be thought of
as biological (fitness as expected relative number of offspring) or cultural (fitness of
likelihood of being imitated or repeated). The types that the dynamic operates on are pairs
consisting of a lexicon and a pragmatic use pattern. A type's expressivity depends on its
communicative efficiency within a population while its learnability depends on the fidelity by
which it is inferred by new generations of learners. The learners' task is consequently to
perform a joint inference over types of linguistic behavior and lexical meaning.

\section{Scalar implicatures}\label{sec:si-case-study}
%
Scalar implicatures are a particularly well-studied type of conventional pragmatic inferences. They are licensed for groups of expressions ordered in terms of informativity, here understood as an entailment induced order. For instance, {\em some} is entailed by {\em all}. If it were true that `Chris owns all of Johnny Cash's albums', it would also be true that `Chris owns some of Johnny Cash's albums'. However, while weaker expressions such as {\em some} are truth-conditionally compatible with stronger alternatives such as {\em all}, this is not what their use is normally taken to convey. Instead, the use of a less informative expression when a more informative one could have been used can license a defeasible inference that stronger alternatives do not hold (cf. \citealt{horn:1972,gazdar:1979}). That is, a hearer who assumes the speaker to be able and willing to provide all relevant information can infer that stronger alternatives do not hold because the speaker used a weaker alternative instead. In this way, `Chris owns some of Johnny Cash's albums' is strengthened to convey that he owns some but not all albums. A bound that rules out stronger alternatives is thusly not codified in the lexical meaning of weak alternatives but instead pragmatically supplied.

@@

\subsection{Analysis}
The dynamics are initialized with an arbitrary distribution over types, constituting a population's first generation. We report the outcome of their development under different parameters and pressures after $50$ generations. This ensures that these outcomes correspond to developmental plateaus after which no noteworthy change is registered. As specified in \S\ref{sec:learnability}, the mutation matrix $Q$ can be obtained by considering all possible state-message sequences of length $k$. Given that this is intractable for large $k$, they were approximated by sampling $250$ sequences from each type's production probabilities.

We begin setting the stage by defining the learning prior and the set of types under consideration. Drawing from our preceding discussion, functional pressure on successful communication combined with learning pressures in the form of a bias against upper-bounds may lead to the selection of $\Llack$-like semantics. However, it is instructive to first inspect the effect of these pressures in isolation. Accordingly, we first showcase the effects that expressivity and learnability have on their own before turning to those resulting from their combination.

\subsubsection{Inductive learning bias}
There is a growing effort to develop empirically  testable  representational  languages that allow for the measure of semantic complexity. For instance, so-called {\em languages of thought} (LOT) have been put to test in various rational probabilistic models that show encouraging results (see, e.g., \citealt{katz+etal:2008, piantadosi+etal:underreview, piantadosi+etal:2012a} and \citealt{piantadosi+jacobs:2016} for recent discussion). At its core, a LOT defines a set of operations and composition rules from which lexical meaning can be derived. As a first approximation and for the sake of concreteness, we follow this approach to make the preference of learners for simpler semantic representations precise by defining the complexity of a concept as a function of its derivation cost in a weighed generative LOT.

>>>>

\begin{abstract}
  According to standard linguistic theory, the meaning of an utterance is the product of 
  conventional semantic meaning and general pragmatic reasoning applied
  to the context of utterance. This implies that models of cultural evolution of meaning
  should likewise take into consideration that observable language use is a complex interaction
  of semantic representations and pragmatic use. To this end, we present a game theoretic model
  of cultural language evolution where communicative pressures work on abstract semantic
  representations and pragmatic patterns of use. Our model traces two evolutionary forces and
  their interaction: (i) fitness-based pressure towards communicative efficiency and (ii)
  systematic transmission perturbations when linguistic knowledge is transferred from one agent
  to another.  We illustrate the model based on a case study showing that learning biases that favor simple semantic
  representations can prevent the lexicalization of pragmatic inferences as long as mutual reasoning counteracts a loss in expressivity otherwise incurred by the adoption of such semantics. 
\end{abstract}

\section{Introduction}\label{sec:introduction}
What is conveyed usually goes beyond what is said. A request for a blanket can be politely
veiled by uttering ``I'm cold.'' The temporal succession of events can be communicated by the
order in which conjuncts appear as in ``I traveled to Paris and got married.'' An invitation
can be declined by saying ``I have to work.'' An influential explanation of the relation
between the literal meaning of expressions and what they are intended and interpreted to convey
is due to \citet{grice:1975}, who characterizes pragmatic use and interpretation as a process of
mutual reasoning about rational language use. For instance, under the assumption that the
speaker is cooperative and relevant, ``I have to work'' may be interpreted as providing a
reason why the speaker will not be able to accept the invitation, going beyond its literal
meaning. Some of these enrichments are rather \emph{ad hoc}. Others show striking regularities,
such as the use of ability questions for polite requests (``Could you please \dots?''), or
certain enrichments of lexical meanings such as \emph{and} to mean \emph{and then}.

A particularly productive and well studied class of pragmatic enrichments are scalar
implicatures
\citep{horn:1984,Hirschberg1985:A-Theory-of-Sca,LevinsonPragmatics1983,Geurts2010:Quantity-Implic}. Usually,
the utterance of a sentence like ``I own some of Johnny Cash's albums'' will be taken to mean
that the speaker does not own all of them. This is because, if the speaker had them all, he
could have used the stronger word \emph{all} instead of \emph{some} in his utterance, 
thereby making a more informative statement. Scalar implicatures, especially the
inference from \emph{some} to \emph{some but not all}, have been studied extensively, both
theoretically
\citep[e.g.,][]{Sauerland2004:Scalar-Implicat,ChierchiaFox2008:The-Grammatical,Rooyvan-RooijJagerde-Jager2012:Explaining-Quan}
as well as experimentally
\citep[e.g.,][]{BottNoveck2004:Some-Utterances,huang+snedeker:2009,GrodnerKlein2010:Some-and-Possib,GoodmanStuhlmuller2013:Knowledge-and-I,DegenTanenhaus2012:Processing-Scal}. While
there is much dispute in this domain about many details, a position endorsed by a
clear majority is that a word like \emph{some} is underspecified to mean \emph{some and maybe
  all} and that the enrichment to \emph{some but not all} is part of some regular 
process with roots in pragmatics.

@@

To fill this gap, we spell out a model of the co-evolution of conventional meaning and
pragmatic reasoning types. The objects of replication and selection are pairs of lexical
meanings and general types of pragmatic behavior, which we represent using state-of-the-art
probabilistic cognitive models of pragmatic language use
\citep{frank+goodman:2012,FrankeJager2015:Probabilistic-p,GoodmanFrank2016:Pragmatic-Langu}. Replication
and selection are described by the \emph{replicator mutator dynamic}, a general and established
model of evolutionary change in large and homogeneous populations
\citep{Hofbauer1985:The-Selection-M,nowak+etal:2000,NowakKomarova2001:Evolution-of-Un,hofbauer+sigmund:2003,Nowak2006:Evolutionary-Dy}. The
approach allows us to study the interaction between (i) evolutionary pressure towards
communicative efficiency and (ii) possible infidelity in the transmission of linguistic
knowledge, caused by factors such as inductive learning biases and sparse data to learn from. Considering
transmission of linguistic knowledge is important because neither semantic meanings nor
pragmatic usage patterns are directly observable. Instead, language learners have to infer these
unobservables from the observable behavior in which they result. We formalize this 
process as a form of Bayesian inference. Our approach thereby contains a well-understood model
of iterated Bayesian learning \citep{griffiths+kalish:2007} as a special case, but combines it
with functional selection, here formalized as the most versatile dynamic from evolutionary game
theory; the replicator dynamic
\citep{TaylorJonker1978:Evolutionary-St}. Section~\ref{sec:model} introduces this model.

@@

An elegant formal approach to capture the interaction between expressivity and learnability
is the \emph{replicator mutator dynamic}
\citep{Hofbauer1985:The-Selection-M,nowak+etal:2000,NowakKomarova2001:Evolution-of-Un,hofbauer+sigmund:2003,Nowak2006:Evolutionary-Dy}. In
its simplest, discrete-time formulation, the RMD defines the frequency $x'_i$ of each type $i$
in a population at the next time step as a function of: (i) the frequency $x_i$ of each type
$i$ before the step, (ii) the fitness $f_i$ of each type $i$ before the step, and (iii) the
probability $Q_{ji}$ that an agent who wants to imitate, adopt, or learn the type of an agent
with type $j$ ends up acquiring type $i$:
\begin{align}
  \label{eq:RMD_discrete}
  x'_i = \sum_j Q_{ji} \frac{x_jf_j}{\sum_k x_k f_k}\,.
\end{align}
The RMD consists of two components: fitness-based selection and transmission biases. This
becomes most transparent when we consider an equivalent formulation in terms of a step-wise
application of the discrete-time replicator dynamic \citep{TaylorJonker1978:Evolutionary-St} on the initial population vector $\vec{x}$
and its subsequent multiplication with a mutation matrix $Q$:

@@

Where our goal is an application of this dynamic to the case of co-evolution of semantic
meaning and pragmatic use, we need to fix what the relevant types are, how fitness is measured
and how the mutation matrix is computed. These issues are addressed, one by one, in the
following.

@@

Lexica codify the truth-conditions of expressions. A convenient way to represent lexica is by
$(\card{\States}, \card{\Messgs})$-Boolean matrices, where $\States$ is a set of states
(meanings) and $M$ a set of messages (forms available in the language). For example, suppose
that there are two relevant world states $\States = \set{\ssome, \sall}$. In state $\ssome$
Chris owns some but not all of Johnny Cash's albums while in $\sall$ Chris owns them
all. Suppose that there are two messages $\Messgs = \set{\msome, \mall}$ where $\msome$ is
short for a sentence like \emph{Chris owns some of Johnny Cash's albums} and $\mall$ for the
same sentence with \emph{some} replaced by \emph{all}.  
Lexica for this case would assign a Boolean truth value, either $0$ or $1$, to each state-message pair. The following two lexica exemplify the distinction
between a lexicalized upper-bound for \emph{some} in $\Lbound$ and the widely assumed logical
semantics with only a lower-bound in $\Llack$.

@@

We distinguish between two kinds of pragmatic behavior. {\em Literal interlocutors} produce and
interpret messages literally, being guided only by their lexica. {\em Pragmatic interlocutors}
instead engage in mutual reasoning to inform their choices. Recent probabilistic models of
rational language use
\citep{franke:2009,frank+goodman:2012,FrankeJager2015:Probabilistic-p,GoodmanFrank2016:Pragmatic-Langu}
capture different types of pragmatic behavior in a reasoning hierarchy. The hierarchy's bottom,
level $0$, corresponds to literal language use. Pragmatic language users of level $n + 1$
act (approximately) rational with respect to level-$n$ behavior of their
interlocutors. (\ref{h:level0}) and (\ref{s:level0}) define probabilistic behavior of literal
hearers and speakers respectively: \mf{is it a problem that $S$ denotes speakers and the set of
states?} \tb{I think that's fine. $S_n(\cdot \mid \cdot)$ is not used beyond this section. Otherwise we can, e.g., relabel $M$ as $F$(orms) and use $M$(eanings) instead of $S$(tates). I'm fine with either option.}
\begin{flalign}
&H_{0}(s \mid m;L) \propto pr(s) L_{sm} \label{h:level0}\\
&S_{0}(m \mid s;L) \propto \exp(\lambda \; L_{sm}) \label{s:level0}
\end{flalign}

According to (\ref{h:level0}), a literal hearer's interpretation of a message $m$ as a state
$s$ depends on her lexicon and her prior over states, $pr \in \Delta(S)$. For simplicity, in
the following this prior is assumed to be uniform. A literal interpreter with lexicon $\Lbound$
from above would assign $\ssome$ a probability of $H_0(\ssome \mid \msome; \Lbound) = 1$ after
hearing $\msome$, while a literal interpreter with lexicon $\Llack$ would assign $\ssome$
probability $H_0(\ssome \mid \msome; \Llack) = 0.5$. As usual in probabilistic pragmatics models, speaker behavior is regulated by a soft-max
parameter $\lambda$, $\lambda \geq 1$ \citep{luce:1959,sutton+barto:1998}. As $\lambda$
increases, choices made in production are more rational in that higher values lead to behavior
that is increasingly in line with expected utility maximization. Expected utility of a message
$\messg$ in state $\state$ for a level $n+1$ speaker is here defined as $H_{n}(s|m;L)$, the probability that the hearer will assign to or choose the correct meaning. Put differently, how likely the hearer is thought to infer $\state$ when sent $\messg$. The formulation
in (\ref{s:level0}) also allows false messages to be sent with a small positive probability in
analogy to the definition of pragmatic speakers in probabilistic pragmatic models. This is also necessary to guarantee a mutation matrix with only positive entries (see below). If $\lambda = 1$ a literal
speaker with either lexicon $\Lbound$ or $\Llack$ produces $\msome$ in $\ssome$ with probability $S_{0}(\msome \mid
\ssome;L_{\text{bound},\text{lack}}) \approx .73$. The probability of producing $\msome$ in $\sall$ is $S_{0}(\msome \mid
\sall;\Lbound) \approx .27$ for $\Lbound$ and $S_{0}(\msome \mid
\sall;\Llack) = .5$ for $\Llack$. By contrast, if $\lambda = 20$, a literal speaker with either lexicon produces $\msome$ in $\ssome$ with probability $S_{0}(\msome \mid
\ssome;L_{\text{bound},\text{lack}}) \approx 1$. But the probability of producing $\msome$ in $\sall$ is $S_{0}(\msome \mid
\sall;\Lbound) \approx 1$ for $\Lbound$ whereas it is still $S_{0}(\msome \mid \sall;\Llack) = .5$ because $\msome$ is semantically associated with both states by $\Llack$.

Pragmatic behavior of level $n+1$ is similar to its literal counterparts but uses the
interpretation or production behavior of a level-$n$ player instead of the lexical meaning:
\begin{flalign}
&H_{n+1}(s|m;L) \propto pr(s) S_{n}(m|s;L) \label{h:leveln}\\
&S_{n+1}(m|s;L) \propto  \exp(\lambda \; H_{n}(s|m;L)) \label{s:leveln}
\end{flalign}

Particularly important to our purpose is the fact that players of a level greater than $0$ using $\Llack$ can come to {\em pragmatically} associate $\msome$ with $\ssome$ over $\sall$ in contrast to their literal counterparts. Intuitively, this is so because hearers reason a speaker to convey $\sall$ with $\mall$, as this has a higher chance of leading to mutual understanding than using underspecified $\msome$. Therefore they will expect $\msome$ to be more strongly associated with $\ssome$ than with $\sall$ as it is already possible to clearly convey the latter state with a different message. The converse reasoning pattern holds for the speaker. In this way, two distinct lexical conventions such as $\Lbound$ and $\Llack$ can give rise to similar observable linguistic behavior.

@@

Languages are shaped not only by functionalist forces towards greater expressivity. Another
important factor is the fidelity by which language is transmitted. Among others, linguistic
production can be prone to errors, states or messages may be perceived incorrectly, and
multiple languages may be compatible with the data learners are exposed to. These sources of
uncertainty introduce variation in the transmission of linguistic knowledge from one generation to the next. In particular, learning biases in the iterated transmission process can influence language evolution
substantially.

@@

Following \citet{griffiths+kalish:2007} we model language acquisition as a process of Bayesian
inference in which learners combine the likelihood of a type producing the witnessed learning
input with prior inductive biases. Experimental and mathematical results on iterated learning
suggest that the outcome of this process reflects learners' inductive biases
\citep[e.g.,][]{kirby+etal:2014}. In a Bayesian setting these biases can be codified in a prior
$P \in \Delta(T)$, which reflects the amount of data a learner requires to faithfully acquire
the language of the teacher \citep[cf.][450]{griffiths+kalish:2007}. The extent of the prior's
influence has been shown to heavily depend on the learning strategy assumed to underlie the
inference process. On the one hand, early simulation results suggested that weak biases could
be magnified by exposing learners to only small data samples \citep[e.g., in][]{brighton:2002}. On the other, \citeposs{griffiths+kalish:2007} mathematical
characterization showed that iterated learning converged to the prior, i.e., that the resulting
distribution over languages corresponds to the learners' prior distribution and is not
influenced by the amount of input given to them. This difference in predictions can be traced
back to differences in the selection of hypotheses from the posterior. Griffith \& Kalish's
convergence to the prior holds for learners that sample from the posterior. More deterministic
strategies such as the adoption of the type with the highest posterior probability, so-called
{\it maximum a posterior estimation} (MAP), increase the influence of both the prior and the
data \citep{griffiths+kalish:2007,kirby+etal:2007}. In the following, we use a parameter
$l\ge1$ to modulate between posterior sampling and the MAP strategy. When $l = 1$ learners
sample from the posterior. The learners' propensity to maximize the posterior grows as $l$
increases.

@@

Expressivity and learnability are central to the cultural evolution of language. These
components can be modelled, respectively, as replication based on a measure of fitness in terms
of communicative efficiency and iterated Bayesian learning. Their interaction is described by
the discrete time replicator mutator dynamics in (\ref{eq:RMD_discrete}), repeated here:
\begin{align*}
  x'_i = \sum_j Q_{ji} \frac{x_jf_j}{\sum_k x_k f_k}\,.
\end{align*}
This equation defines the frequency $x'_i$ of type $i$ at the next time step, based on its
frequency $x_i$ before the step, its fitness $f_i$ and the probability that a learner infers
$i$ when observing the behavior of a type-$j$ agent. Fitness-based selection can be thought of
as biological (fitness as expected relative number of offspring) or cultural (fitness as 
likelihood of being imitated or repeated). The types that the dynamic operates on are pairs
consisting of a lexicon and a pragmatic use pattern. A type's expressivity depends on its
communicative efficiency within a population while its learnability depends on the fidelity by
which it is inferred by new generations of learners. The learners' task is consequently to
perform a joint inference over types of linguistic behavior and lexical meaning.

\section{Scalar implicatures}\label{sec:si-case-study}
%
Scalar implicatures are a particularly well-studied type of conventional pragmatic inferences. They are licensed for groups of expressions ordered in terms of informativity, here understood as an entailment induced order. For instance, {\em some} is entailed by {\em all}. If it were true that ``Chris owns all of Johnny Cash's albums'', it would also be true that ``Chris owns some of Johnny Cash's albums''. However, while weaker expressions such as {\em some} are truth-conditionally compatible with stronger alternatives such as {\em all}, this is not what their use is normally taken to convey. Instead, the use of a less informative expression when a more informative one could have been used can license a defeasible inference that stronger alternatives do not hold (cf. \citealt{horn:1972,gazdar:1979}). That is, a hearer who assumes the speaker to be able and willing to provide all relevant information can infer that stronger alternatives do not hold because the speaker used a weaker alternative instead. In this way, ``Chris owns some of Johnny Cash's albums'' is strengthened to convey that he owns some but not all albums. A bound that rules out stronger alternatives is thusly not codified in the lexical meaning of weak alternatives but instead pragmatically supplied.

@@

\subsection{Analysis}
The dynamics are initialized with an arbitrary distribution over types, constituting a population's first generation. In the following we report the outcome of their development under different parameters and pressures after $50$ generations. This ensures that these outcomes correspond to developmental plateaus after which no noteworthy change is registered. As specified in \S\ref{sec:learnability}, the mutation matrix $Q$ can be obtained by considering all possible state-message sequences of length $k$. Given that this is intractable for large $k$, the sets of data learners are exposed to are approximated by sampling $250$ $k$-length sequences from each type's production probabilities.

We begin setting the stage by defining the learning prior and the set of types under consideration. Drawing from our preceding discussion, functional pressure on successful communication combined with learning pressures in the form of a bias against upper-bounds may lead to the selection of $\Llack$-like semantics. However, it is instructive to first inspect the effect of these pressures in isolation. We therefore first showcase the effects that expressivity and learnability have on their own before turning to those resulting from their combination.

\subsubsection{Inductive learning bias against semantic complexity}
There is a growing effort to develop empirically  testable  representational  languages that allow for the measure of semantic complexity. For instance, so-called {\em languages of thought} (LOT) have been put to test in various rational probabilistic models that show encouraging results (see, e.g., \citealt{katz+etal:2008, piantadosi+etal:underreview, piantadosi+etal:2012a} and \citealt{piantadosi+jacobs:2016} for recent discussion). At its core, a LOT defines a set of operations and composition rules from which lexical meaning can be derived. As a first approximation and for the sake of concreteness, we follow this approach to make the preference of learners for simpler semantic representations precise and define the complexity of a representation as a function of its derivation cost in a weighed generative LOT.

Our toy grammar of concepts is given in Table~\ref{tab:grammar}. This grammar uses basic set-theoretic operations to form expressions which can be evaluated as true or false in different world states such as $\ssome$ or $\sall$ (see below). Applications of generative  rules have a cost attached to them. (Alternatively, a probability.) Here we simply assume that Boolean combinations of concepts are more complex than atomic concepts and that otherwise each rule application adds the same cost unit.
\begin{table}
  \centering
  \begin{align*}
    C & \rightarrow_2 C \wedge C 
    & 
    X & \rightarrow_1 \set{A,B} \\
    C & \rightarrow_2 \neg C 
    & 
    X & \rightarrow_1 X \cap X \\
    C & \rightarrow_1 X \subseteq X
    & 
    X & \rightarrow_1 X \cup X \\
    C & \rightarrow_1 X \neq \emptyset \\
    C & \rightarrow_1 X = \emptyset     
  \end{align*}
  \caption{Toy grammar in a set-theoretic LOT with weighted rules.}
  \label{tab:grammar}
\end{table}

The prior probability of a type is just the prior probability of its lexicon. The prior of a lexicon is a function of the complexity of the lexical representations in its image set. As motivated above, lexica that use simpler concepts are \emph{a priori} more likely. A simple way of defining such priors over lexica is:
\begin{align*}
  P(L)  & \propto \prod_{c \in Im(L)} P(c)   \ \text{, with} & 
  P(c) & \propto \max_{c'}Compl(c') - Compl(c) + 1\,,
\end{align*}
